### 2023-06-09 Manuscript/Draft1.Rmd
Manuscript update including representative plots

We construct spatial models for Yellowtail Flounder, Atlantic Cod, and Haddock on Georges Bank, where movement of each species combine static species-specific habitat preferences with biologically-based temperature preferences. Model dynamics are driven by a time series of temperature gradients to create simulated data sets for each population where the true biomass is known. Using temperature gradients that repeat each year creates data sets with repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time. We conduct stratified random sampling on our simulation output to mimic the bottom trawl survey and use the samples to compare the ability of contemporary indexing methods to track population trends.

@@

Each species is assumed to have normally distributed temperature preferences $N(\mu,\sigma)$, with mean $\mu$ and standard deviation $\sigma$. Values were chosen by combining information in the literature with temperatures recorded in the bottom trawl survey. We assume Yellowtail Flounder's preferences are $N(8.75,4.25)$, while Haddock and Cod have preferences $N(9,4)$. Weekly estimated temperature data for the region for 2012 was obtained from FVCOM [@chen2006unstructured]. We chose to repeat temperature estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data. The 2012 data was chosen because it displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values, allowing for a smooth repeating yearly temperature pattern for the constant temperature scenario. The 2012 temperature data was also transformed to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation. We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on dynamics while remaining within reasonable computational limits in terms of the length of the simulation. Figure \@ref(fig:temp-scenarios) depicts mean trends for the temperature scenarios used in our models. **dont forget to include animated gif in final submission**

@@

We carry out 20 year simulations for each of our three species under various population scenarios. Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year [[can cite the 2022 management track assessments, see https://apps-nefsc.fisheries.noaa.gov/saw/sari.php for when the document becomes available]] **cite**. For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock. To provide a comprehensive analysis of population indexing methods we consider all possible population trend combinations for Yellowtail Flounder. The specific population trends used in our analyses can be see in Figure \@ref(fig:pop-scenarios). Each of these scenarios is simulated twice: first with with an oscillating temperature gradient that repeats and second with a temperature gradient that increases roughly 5 degrees Celsius over the duration of the 20 year simulation, for a total of 10 simulated spatial datasets.

@@

The top two panels in Figure \@ref(fig:PopPct) depict the results for Yellowtail Flounder with a repeating temperature gradient on the left (constant temperature) and a temperature gradient that increases over time on the right (increasing temperature). In both temperature scenarios we see the percent of Yellowtail Flounder in strata 13 decrease over the course of the time series in both seasons. The spring population in stratum 19 also decreases in both temperature scenarios as well. The percent of the population increases in stratum 16 in the spring over the duration of both time series, which implies the flow out of strata 13 and 19 in the spring are going into stratum 16. These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail Flounder that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12. These spring changes are therefore related to the static habitat values in our model rather than the temperature preferences, which is why we seen the same dynamics with constant and increasing temperature. While we observe similar changes in the fall (weeks 37 and 38) in the constant temperature scenario, an increasing average temperature results in a decrease in the population in stratum 16 over time and corresponding increases in strata 17 and 18 (see Figure \@ref(fig:PopPct)). These dynamics imply that an increase in temperature results in the more shallow strata 16 becoming less desirable than the deeper and more narrow outer strata 17 and 18. One noticeable seasonal difference in the constant temperature scenario for Yellowtail is how ~10\% of the population exists in the narrow outer strata 17 in the fall, while seemingly none of the population exists in any of the strata near the edge of the domain (14, 15, 17, 18) in the spring.

Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata) contain the absolute error between our abundance estimates for Yellowtail Flounder comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B). In reviewing these Tables we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors. The settings used in VAST mattered with Settings B outperforming Settings A in $67/96\approx 70\%$ of scenarios. When all strata are sampled, adding covariates improved estimates in all scenarios. Including covariates still improve VAST estimates when certain strata are excluded from sampling, but the change in error was smaller than with all strata included. There are several instances in which VAST failed to provide improved abundance estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata), but including covariate information to each VAST model produced estimates with significantly lower error than the corresponding stratified mean estimate

Of all scenarios without covariates, $33/48\approx68$\% had a VAST fit with a lower relative error compared to the stratified mean estimate. All 15 of the covariate-free VAST estimates that resulted in a higher error than the stratified mean were for the fall season and had a common theme of producing abundance estimates that are above the true model value. These 15 fall estimates span all other scenario variations. The implication of this is that our model-based approach without covariates struggles with the primary seasonal difference for Yellowtail Flounder, which is that a larger percentage of the population exists in the narrow strata toward the edge of the domain (18 and/or 17). This theory is further supported by the fact that the absolute error in the increasing temperature scenarios increased dramatically in the fall season, when the combined percentage of the population in the outer strata 17 and 18 increased to over 40\% by the end of the simulation.

Including covariate information made a noticeable difference in our Yellowtail Flounder model-based VAST estimates. All of the VAST estimates that included covariate information produced a lower relative error than the corresponding VAST model that did not include covariates. The largest improvements were seen in the increasing temperature scenarios. When comparing to the design-based estimates, $47/48\approx98$\% of the VAST estimates that included covariates information had a lower relative error than the corresponding stratified mean estimate. This implies that the covariate information helped our model-based estimate account for the design-based issues related to an increasing percentage of the population entering smaller strata, which can become exacerbated in the increasing temperature simulations.

We see diminished performance of our abundance estimates for Yellowtail under increasing temperature, with the most dramatic changes seen in our VAST estimates without covariates. One exception to this is when we use a stratified mean approach while the population is decreasing. Our analyses have found that the stratified mean tends to under estimate the true abundance and since these estimates are bounded below by zero, as the Yellowtail population decreases towards zero the difference between the estimate and the true value also decrease. That is, if the population is low enough, failing to appropriately sample the population in a design-based method produces the same result as appropriately sampling. In comparing abundances estimates calculated under the same conditions with the only difference being the temperature scenario, we see that an increasing average temperature had a larger impact on the model-based estimates compared to the design-based method. Specifically, when comparing the error of abundance estimates from the constant temperature scenario to the corresponding increasing temperature scenario, VAST increased by an average factor of 1.75 (both with and without covariates) while the  stratified mean error increased by an average factor of 1.25.

@@

In the constant temperature scenario for Cod, the population decreases its presence in strata 19 and 20 in both seasons over the duration of the simulation, while simultaneously increasing presence in stratum 16. However, we see a seasonal impact in stratum 16 during the increasing average temperature simulations where in the fall the population decrease presence in strata 16 and 21, and increase presence in 18, 22, and 24 (see Figures \@ref(fig:strata-plot) and \@ref(fig:PopPct)). Similar to the Yellowtail Flounder population, the favorable habitat in stratum 16 acts as an attractor in both temperature scenarios in the spring when the water temperature is cooler. When the temperature increases over time, the fall population compensates by shifting their preference to the adjacent strata that are deeper and/or further north than stratum 16.

Table \@ref(tab:CodERROR) contains the absolute error between our abundance estimates for Cod and the true model values. The VAST different VAST settings produced similar error values with $18/32\approx57%$ of estimates performing better with VAST Settings A. Of the abundance estimates without covariates, $12/16=75$\% had a VAST fit with a lower relative error compared to the stratified mean estimate. VAST produced higher relative error compared to the stratified mean in scenarios that involved increasing temperature in the fall season without covariates. Similar to the Yellowtail results, this implies the model-based approach had seasonal trouble with populations shifting into smaller strata where fewer samples take place. Providing covariate information in these cases once again helped the model-based approach to provide improved absolute error estimates relative to the stratified mean. However, we see that adding covariates to VAST in the fall produces higher absolute error values when sampling reduced strata in the constant temperature scenario.

@@

<!-- Similar to Yellowtail Flounder results, Cod are shifting their spatial distribution into the smaller exterior strata in these conditions, which is decreasing the accuracy of the covariate-free VAST estimate. While adding covariates to these fits allowed for an improved error relative to the stratified mean, the VAST abundance still display an overestimate early in the time series before covariates correct the trend, allowing for a lower overall absolute error **(show these plots? IncTemp fall scenarios)**. -->

@@

Table \@ref(tab:HadERROR) contains the absolute error between our abundance estimates for Haddock and the true model values. Absolute error values were lower with VAST Settings B in $21/32\approx 66\%$ of scenarios. We notice that the model-based VAST produced particularly large errors in spring compared to the stratified mean, with added covariates only improving to the level of the stratified mean. VAST shows improved results in fall relative to the stratified mean with incluidng covariates producing extremely low errors in some cases. For Haddock results, adding covariates improves estimates only when all strata are included. That is, similar to Cod, when sampling a reduced domain adding covariates actually decreases VAST's accuracy. Since this occurs in all scenarios, it seems to again be related to failing to accurately monitor stratum 17 near the edge of the domain.

@@

Our analysis showed that both the design-based and model-based approaches are each capable of providing biomass estimates that track the true abundance trend. While VAST estimates tended to provide lower relative error values compared to the stratified mean, especially when covariate information was included in the models, VAST Settings A and B combined to produce over 30 abundance estimates with larger error values than the largest stratified mean error. These facts highlight the power of model-based estimates as well as how sensitive they are to the choices made regarding settings and data used in analyses. This problem can be compounded by the lack of clear guidance regarding how to build and analyze models with the correct inputs. Even more concerning, a recent simulation study showed that the existing diagnostic tools available in VAST sometimes guide users towards using settings that decrease the accuracy of the model **cite Chris C paper**. The value of model-based approaches such as VAST are clear- they allow users to tailor the model to overcome issues related to the input data by and to account for factors such as environmental change. However, the many consequential decisions that must be made in a given model and the lack of clear criteria for making these decisions makes it challenging to make these choices and even more difficult to determine whether the choices are correct.

Spatial factors contributed to many of the scenarios that produced larger error values seen in our estimates. Flow from the shallow centrally located strata into the more deeper, smaller, and less-sampled eastern strata contributed to VAST failing to accurately modeling the abundance of yellowtail flounder without the addition of covariate information, the impact of which was greatly exacerbated in the increasing temperature scenarios. While adding covariates can help inform the model and improve the estimate as seen with the yellowtail flounder simulation results, including covariates can hinder the model in certain instances. For example, including covariates for haddock when the northern most strata were removed from the survey meant that the population was not being sampled in the coldest part of its domain which led to an incorrect covariate response that ultimately increased the error between the VAST estimate and the true model abundance.

Seasonal differences in spatial abundance patterns in our population models led to seasonal differences in abundance estimates, with most estimates producing a higher error in the fall.  
spawn weeks 9-14
something about seasonal impact seen in all estimates with fall estimates typically noticeably worse. We explored this through the ratio plots. This was mostly a problem with VAST as SM showed more consistent estimates between seasons.

<!-- <font size="5">*Impact of Covariates*</font> -->

As seen in Table **cite VAST setting table**, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses. Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with about 73% had a lower absolute error compared to the corresponding estimate without covariates ($117/160\approx 73\%$), and $77/80\approx 96\%$ of VAST models with covariates provided lower absolute error than the corresponding stratified mean estimate. There is more to explore with respect to the impact of covariates. For example, by including just one of the covariates individually, one could test which had a larger impact on abundance estimates. Adding noise to our covariate input would test how robust the model-based estimates are to uncertainty in the covariate information. One could also test the impact of assuming the wrong covariate response function (linear vs polynomial etc). 
<!-- **For discussion: clearly our perfect covariate information made a big difference for yellowtail flounder (and others), but the question is could we achieve similar results without "perfect" information? Which covariate is best? Good question for future research** -->

 Of the ~27% of VAST models where including covariates did not improve the VAST estimate, most provided a comparable error (for ex, 0.11 vs 0.13). The instances where including covariates provided a significant different in error took place in scenarios that included either an increasing temperature and/or reduced sampling domain. More specifically, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates. However, when a reduced number of strata are sampled for Haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates. The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, leading to an incomplete estimation of the covariate response. Figure \@ref(fig:strata-plot) shows the strata excluded from sampling and Figure \@ref(fig:CovResponse) depicts the resulting covariate response for Yellowtail Flounder and Haddock for the same population scenario. Although strata were excluded for both species, the survey samples for Yellowtail Flounder still ranged across the species preferred values and thus fully formed the covariate response, while the Haddock covariate response did not contain enough samples from the lower temperature range due to the specific strata that were excluded from sampling (ie, northern strata). As a result, when projecting into those strata as seen in the lower portion of Figure \@ref(fig:CovResponse), Yellowtail estimates with covariate information approximate the correct trend despite these strata being excluded from sampling, while the Haddock estimates that include strata provide that wrong trend and estimate the wrong magnitude of biomass.

@@

\section{Acknowledgements}
[[we should thank Jim (obviously), but also those who helped you with the habitat data (robyn, david Chev.) and others? also should note that funding provided by Northeast Fisheries and Climate program at the Northeast Fisheries Science Center ]]  Also thank Joe Langan for helping develop the proposal that funded this work.

@@

```{r CovResponse, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Temperature covariate response plots and resulting population estimate for Yellowtail Flounder on the left and Haddock on the right. In both cases the spatial simulations that were surveyed had an increasing population over time, increasing average temperature, and certain strata excluded from sampling, as shown in Figure 1**check**.",out.width="90%"}
# knitr::include_graphics(c("Images/covariate_good.png","Images/covariate_bad.png"))
# knitr::include_graphics(c("Images/covariate_est.png"))

knitr::include_graphics(c("Images/covariate_plot.png"))

>>>>

We construct spatial models for yellowtail flounder, Atlantic Cod, and Haddock on Georges Bank, where movement of each species combine static species-specific habitat preferences with biologically-based temperature preferences. Model dynamics are driven by a time series of temperature gradients to create simulated data sets for each population where the true biomass is known. Using temperature gradients that repeat each year creates data sets with repeating spatial patterns, whereas using a temperature gradient that increases on average throughout the simulation leads to spatial distributions that shift over time. We conduct stratified random sampling on our simulation output to mimic the bottom trawl survey and use the samples to compare the ability of contemporary indexing methods to track population trends.

@@

Each species is assumed to have normally distributed temperature preferences $N(\mu,\sigma)$, with mean $\mu$ and standard deviation $\sigma$. Values were chosen by combining information in the literature with temperatures recorded in the bottom trawl survey. We assume yellowtail flounder's preferences are $N(8.75,4.25)$, while Haddock and Cod have preferences $N(9,4)$. Weekly estimated temperature data for the region for 2012 was obtained from FVCOM [@chen2006unstructured]. We chose to repeat temperature estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data. The 2012 data was chosen because it displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values, allowing for a smooth repeating yearly temperature pattern for the constant temperature scenario. The 2012 temperature data was also transformed to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation. We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on dynamics while remaining within reasonable computational limits in terms of the length of the simulation. Figure \@ref(fig:temp-scenarios) depicts mean trends for the temperature scenarios used in our models. **dont forget to include animated gif in final submission**

@@

We carry out 20 year simulations for each of our three species under various population scenarios. Historically, Atlantic Cod has seen significant decline over the last XXX years while Haddock has increased in abundance in recent year [[can cite the 2022 management track assessments, see https://apps-nefsc.fisheries.noaa.gov/saw/sari.php for when the document becomes available]] **cite**. For this reason we compare indexing estimates using stratified random samples from decreasing population scenarios for Cod and increasing population scenarios for Haddock. To provide a comprehensive analysis of population indexing methods we consider all possible population trend combinations for yellowtail flounder. The specific population trends used in our analyses can be see in Figure \@ref(fig:pop-scenarios). Each of these scenarios is simulated twice: first with with an oscillating temperature gradient that repeats and second with a temperature gradient that increases roughly 5 degrees Celsius over the duration of the 20 year simulation, for a total of 10 simulated spatial datasets.

@@

The top two panels in Figure \@ref(fig:PopPct) depict the results for yellowtail flounder with a repeating temperature gradient on the left (constant temperature) and a temperature gradient that increases over time on the right (increasing temperature). In both temperature scenarios we see the percent of yellowtail flounder in strata 13 decrease over the course of the time series in both seasons. The spring population in stratum 19 also decreases in both temperature scenarios as well. The percent of the population increases in stratum 16 in the spring over the duration of both time series, which implies the flow out of strata 13 and 19 in the spring are going into stratum 16. These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for yellowtail flounder that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12. These spring changes are therefore related to the static habitat values in our model rather than the temperature preferences, which is why we seen the same dynamics with constant and increasing temperature. While we observe similar changes in the fall (weeks 37 and 38) in the constant temperature scenario, an increasing average temperature results in a decrease in the population in stratum 16 over time and corresponding increases in strata 17 and 18 (see Figure \@ref(fig:PopPct)). These dynamics imply that an increase in temperature results in the more shallow strata 16 becoming less desirable than the deeper and more narrow outer strata 17 and 18. One noticeable seasonal difference in the constant temperature scenario for Yellowtail is how ~10\% of the population exists in the narrow outer strata 17 in the fall, while seemingly none of the population exists in any of the strata near the edge of the domain (14, 15, 17, 18) in the spring.

Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata) contain the absolute error between our abundance estimates for yellowtail flounder comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B). In reviewing these Tables we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors. The settings used in VAST mattered with Settings B outperforming Settings A in $67/96\approx 70\%$ of scenarios. When all strata are sampled, adding covariates improved estimates in all scenarios. Including covariates still improve VAST estimates when certain strata are excluded from sampling, but the change in error was smaller than with all strata included. There are several instances in which VAST failed to provide improved abundance estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata), but including covariate information to each VAST model produced estimates with significantly lower error than the corresponding stratified mean estimate.

Of all scenarios without covariates, $33/48\approx68$\% had a VAST fit with a lower relative error compared to the stratified mean estimate. All 15 of the covariate-free VAST estimates that resulted in a higher error than the stratified mean were for the fall season and had a common theme of producing abundance estimates that are above the true model value. These 15 fall estimates span all other scenario variations. The implication of this is that our model-based approach without covariates struggles with the primary seasonal difference for yellowtail flounder, which is that a larger percentage of the population exists in the narrow strata toward the edge of the domain (18 and/or 17). This theory is further supported by the fact that the absolute error in the increasing temperature scenarios increased dramatically in the fall season, when the combined percentage of the population in the outer strata 17 and 18 increased to over 40\% by the end of the simulation.

Including covariate information made a noticeable difference in our yellowtail flounder model-based VAST estimates. All of the VAST estimates that included covariate information produced a lower relative error than the corresponding VAST model that did not include covariates. The largest improvements were seen in the increasing temperature scenarios. When comparing to the design-based estimates, $47/48\approx98$\% of the VAST estimates that included covariates information had a lower relative error than the corresponding stratified mean estimate. This implies that the covariate information helped our model-based estimate account for the design-based issues related to an increasing percentage of the population entering smaller strata, which can become exacerbated in the increasing temperature simulations.

We see diminished performance of our abundance estimates for Yellowtail under increasing temperature, with the most dramatic changes seen in our VAST estimates without covariates. One exception to this is when we use a stratified mean approach while the population is decreasing. A representative example of yellowtail abundance estimates is shown in Figure \@ref(fig:YTEstEx), where one can visualize the impact on biomass estimates between constant and increasing temperature, and how including covariates can improve the estimate. Our analyses have found that the stratified mean tends to under estimate the true abundance and since these estimates are bounded below by zero, as the Yellowtail population decreases towards zero the difference between the estimate and the true value also decrease. That is, if the population is low enough, failing to appropriately sample the population in a design-based method produces the same result as appropriately sampling. In comparing abundances estimates calculated under the same conditions with the only difference being the temperature scenario, we see that an increasing average temperature had a larger impact on the model-based estimates compared to the design-based method. Specifically, when comparing the error of abundance estimates from the constant temperature scenario to the corresponding increasing temperature scenario, VAST increased by an average factor of 1.75 (both with and without covariates) while the  stratified mean error increased by an average factor of 1.25.

@@

In the constant temperature scenario for Cod, the population decreases its presence in strata 19 and 20 in both seasons over the duration of the simulation, while simultaneously increasing presence in stratum 16. However, we see a seasonal impact in stratum 16 during the increasing average temperature simulations where in the fall the population decrease presence in strata 16 and 21, and increase presence in 18, 22, and 24 (see Figures \@ref(fig:strata-plot) and \@ref(fig:PopPct)). Similar to the yellowtail flounder population, the favorable habitat in stratum 16 acts as an attractor in both temperature scenarios in the spring when the water temperature is cooler. When the temperature increases over time, the fall population compensates by shifting their preference to the adjacent strata that are deeper and/or further north than stratum 16.

Table \@ref(tab:CodERROR) contains the absolute error between our abundance estimates for Cod and the true model values. The different VAST settings produced similar error values with $18/32\approx57\%$ of estimates performing better with VAST Settings A. Of the abundance estimates without covariates, $12/16=75$\% had a VAST fit with a lower relative error compared to the stratified mean estimate. VAST produced higher relative error compared to the stratified mean in scenarios that involved increasing temperature in the fall season without covariates. Similar to the Yellowtail results, this implies the model-based approach had seasonal trouble with populations shifting into smaller strata where fewer samples take place. These impacts are shown in the plots of representative biomass estimates shown in Figure \@ref(fig:CodEstEx). Providing covariate information in these cases once again helped the model-based approach to provide improved absolute error estimates relative to the stratified mean. However, we see that adding covariates to VAST in the fall produces higher absolute error values when sampling reduced strata in the constant temperature scenario.

@@

<!-- Similar to yellowtail flounder results, Cod are shifting their spatial distribution into the smaller exterior strata in these conditions, which is decreasing the accuracy of the covariate-free VAST estimate. While adding covariates to these fits allowed for an improved error relative to the stratified mean, the VAST abundance still display an overestimate early in the time series before covariates correct the trend, allowing for a lower overall absolute error **(show these plots? IncTemp fall scenarios)**. -->

@@

Table \@ref(tab:HadERROR) contains the absolute error between our abundance estimates for haddock and the true model values. Absolute error values were lower with VAST Settings B in $21/32\approx 66\%$ of scenarios. We notice that the model-based VAST produced particularly large errors in spring compared to the stratified mean, with added covariates only improving to the level of the stratified mean. VAST shows improved results in fall relative to the stratified mean with including covariates producing extremely low errors in some cases. For haddock results, adding covariates improves estimates only when all strata are included. That is, similar to Cod, when sampling a reduced domain adding covariates actually decreases VAST's accuracy. Since this occurs in all scenarios, it seems to again be related to failing to accurately monitor stratum 17 near the edge of the domain. Some representative biomass estimates for Haddock that demonstrate these results can be seen in Figure \@ref(fig:HadEstEx).

@@

Our analysis showed that both the design-based and model-based approaches are each capable of providing biomass estimates that track the true abundance trend. While VAST estimates tended to provide lower relative error values compared to the stratified mean, especially when covariate information was included in the models, VAST Settings A and B also combined to produce over 30 abundance estimates with larger error values than the largest stratified mean error. These facts demonstrate the power of model-based estimates while also highlighting how sensitive they are to the choices made regarding settings and data used in analyses. Users often rely on traditional model-building processes and diagnostic techniques such as AIC to guide their choices, but successfully navigating the many decisions depends on the background of the given user and care should be taken to which diagnostic tools are used. For example, a recent simulation study showed that the existing diagnostic tools available in VAST sometimes guide users towards using settings that decrease the accuracy of the model **cite Chris C paper**. The value of model-based approaches such as VAST are clear- they allow users to overcome issues related to the input data by and to account for factors such as environmental change. However, the many consequential decisions that must be made in a given model and the lack of clear criteria for making these decisions makes it challenging to make these choices and even more difficult to determine whether the choices are correct.

Seasonal differences in spatial abundance patterns in our population models led to seasonal differences in abundance estimates, with most estimates producing a higher error in the fall. This may have to do with the fact that our species spawn during months 9-14, which coincides with the spring sampling weeks 13 & 14. Spawning concentrates much of each population in the spawning grounds towards the center of the domain where more samples occur. The populations are more spread out during the fall survey in weeks 37 & 38 and since the water temperature is warmer during these months biomass shifts towards the deeper strata, which tend to be less sampled. We explored this through abundance estimate ratios which revealed how the model-based estimates had a larger change in average ratio between seasonal estimates compared to the design-based estimates.

Spatial factors contributed to many of the scenarios that produced larger error values seen in our estimates. Flow from the shallow centrally located strata into the deeper, smaller, and less-sampled eastern strata contributed to VAST failing to accurately modeling the abundance of yellowtail flounder without the addition of covariate information, the impact of which was exacerbated in the increasing temperature scenarios. While adding covariates can help inform the model and improve the estimate as seen with the yellowtail flounder simulation results, covariates can hinder the model in certain instances.

<!-- <font size="5">*Impact of Covariates*</font> -->
<!-- ($117/160\approx 73\%$) -->
<!-- $77/80\approx 96\%$ -->

As seen in Table 3, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses. Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with about 73% had a lower absolute error compared to the corresponding estimate without covariates , and $\approx 96\, and $77/80\approx 96\%$ of VAST models with covariates provided lower absolute error than the corresponding stratified mean estimate. There is more to explore with respect to the impact of covariates. For example, by including just one of the covariates individually, one could test which had a larger impact on abundance estimates. Adding noise to our covariate input would test how robust the model-based estimates are to uncertainty in the covariate information. One could also test the impact of assuming the wrong covariate response function (linear vs polynomial etc). 
<!-- **For discussion: clearly our perfect covariate information made a big difference for yellowtail flounder (and others), but the question is could we achieve similar results without "perfect" information? Which covariate is best? Good question for future research** -->
As seen in Table 3, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses. Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with about 73% had a lower absolute error compared to the corresponding estimate without covariates As seen in Table 3, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses. Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with about 73% had a lower absolute error compared to the corresponding estimate without covariates 
Of the ~27% of VAST models where including covariates did not improve the VAST estimate, most provided a comparable error (for ex, 0.11 vs 0.13). The instances where including covariates provided a noticeable change in error took place in scenarios that included either an increasing temperature and/or reduced sampling domain. More specifically, when a reduced number of strata are sampled for haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates. This is in contrast to our yellowtail flounder results, where adding covariate information to VAST models always decreased the absolute error in the resulting abundance estimates.  Although strata were excluded for both species, the decline in performance for just Cod and haddock can be explained by a failure to sample the full spectrum of temperature values where the species exists. We notice in Figure \@ref(fig:strata-plot) that the strata excluded from sampling for yellowtail flounder still allowed the survey to range across the species preferred values and thus provided a covariate response that improved the estimate. On the other hand, the specific strata that were excluded from sampling for haddock were located in the northern region of the domain, which means the covariate response for haddock was not informed by any data from the coldest temperature range. This could explain the failure to form a complete covariate response that not only degrades the estimate in the regions being sampled, but also  confounds the ability to project estimates into an unsampled region. Figure \@ref(fig:CovResponse) depicts the impact on biomass estimates of including a covariate response for Haddock when both the average temperature and biomass were increasing over time. When projecting into the unsampled strata as seen Figure \@ref(fig:CovResponse), projecting biomass into the unsampled strata provides the wrong magnitude of biomass and also a different trend.

@@

\section{Acknowledgements}
The authors would like to thank Joe Langan for helping to develop the proposal that funded this work. Funding was provided by the Northeast Fisheries and Climate program at the Northeast Fisheries Science Center. Thank you to Jim Thorson for consulting with us about VAST settings and implementation. We appreciate the help from Robyn Linner for helping to extract FVCOM temperature data and are thankful for David Chevrier for meeting with us to clarify questions related to some of our GIS data.

@@

```{r YTEstEx, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Representative example of bioamass estimate trend for yellowtail flounder. Both scenarios have a relatively constant population with all strata being surveyed. The main difference is the left column has a repeating temperature gradient while the right column has increasing average temperature. ",out.width="90%"}

knitr::include_graphics(c("Images/Yellowtail_Estimate_Example.PNG"))

```{r CodEstEx, echo=FALSE,out.width="100%",fig.show='hold', fig.align='center', fig.cap="Representative example of bioamass estimate trend for atlantic cod. Both scenarios have a decreasing population trend with all strata being surveyed. The main difference is the left column has a repeating temperature gradient while the right column has increasing average temperature",out.width="90%"}

knitr::include_graphics(c("Images/Cod_Estimate_Example.PNG"))

```{r HadEstEx, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Representative example of bioamass estimate trend for haddock. Both scenarios have an increasing population trend and a repeating temperature gradient. The main difference is the left column has all strata being surveyed while the right column has certain strata excluded from sampling",out.width="90%"}

knitr::include_graphics(c("Images/Haddock_Estimate_Example.PNG"))

```{r CovResponse, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Temperature covariate response plot and resulting population estimate for Haddock when the population was increasing over time, the average tempearture was increasing, and certain strata shown in Figure 1 were excluded from sampling.",out.width="90%"}
# knitr::include_graphics(c("Images/covariate_good.png","Images/covariate_bad.png"))
# knitr::include_graphics(c("Images/covariate_est.png"))

knitr::include_graphics(c("Images/covariate_plot3.png"))
